Automatic Matching of Bullet Land Impressions

### 2017-06-28 imaging-paper.tex
text size increased
Resubmission again!

We work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in forensic science. We have developed an open-source package for analyzing bullet lands written in R \citep{R}. This package is called bulletr \citep{bulletr} and enables direct reading and manipulation of x3p files. It also implements all of the methods we propose in this paper. A different package exists for reading x3p files called x3pr \citep{x3pr}, developed by Petraco (2014), but it is not designed to carry out calculations like the ones we propose after the x3p files have been read.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[t]{\textwidth}\centering
\caption{View of the data along the circumference of the bullet (circular segment of about 30 degrees).\label{fig:sidex3p}}{%
      \includegraphics[width=.65\textwidth]{images/sidex3p.png}
      \hspace{1cm}
      \includegraphics[width=.25\textwidth]{images/side-sketch.png}
    }
\end{subfigure}    
\begin{subfigure}[t]{\textwidth}\centering
    \caption{Frontal view of a bullet land (lower end of the view is the bottom of the bullet). \label{fig:topx3p}}{%
    \includegraphics[width=.65\textwidth]{images/topx3p.png} \hspace{1cm}
    \includegraphics[width=.15\textwidth]{images/top-sketch.png}
    }
\end{subfigure}
\caption{Example of a groove-to-groove scan of a bullet land. The red-dotted rectangle on the right shows the location and orientation of the segment. }
\label{fig:allx3p}
\end{figure}

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent groove impressions on the bullet, called a ``land" or land impression. The shoulders are the raised portion of the surface between the groove impression and the land impression. In the Hamby data, typical length (shoulder-to-shoulder) of a land is about 1998.28 micrometers or 2 millimeters. For notational simplicity, we refer to a particular land of a bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

\end{knitrout}
\caption{\label{fig:fixedX}Side profile of the surface measurements (in~$\mu m$) of a bullet land at a fixed height $x$. Note that the global features dominate any deviations, corresponding to the individual characteristics of striation marks.}
\end{figure}

@@

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities between signatures. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/crosscuts-1}

@@

\subsection{Impact of bullet height}\label{sec:heights}
The height at which signatures are extracted for a comparison between bullet lands matters -- signatures taken from heights that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity for quality control: we have to be aware of the height that was used in a matching. Visually, matches degrade if the signatures upon which the match is based are from heights further than 200$\mu m$ apart  (see Supplement Section~\ref{supp:ccf} for more discussion). However, we can extract signatures from multiple heights of the same bullet land for an initial assessment of its quality. By comparing signatures from heights that are not too far apart --  25~$\mu m$ to 50~$\mu m$ -- we get an indication whether the signatures come from a rapidly changing section of the surface, indicative of a break-off or some other damage, or from a stable section, where we have a reasonable expectation of finding matches to other signatures. In the approach here, we keep increasing the height $x$ at which the signature is taken until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet~1-1 from barrel 3, where `stability' is  defined as two aligned signatures from heights chosen 25$\mu m$ apart having a cross-correlation of at least 0.95.

@@

\end{knitrout}
\caption{\label{fig:crosscuts2}Signatures  for barrel 3, bullet~1-1 extracted from varying heights. Initially, the match between signatures taken at heights 25$\mu m$ apart is affected strongly by some break off at the bottom of the bullet. At a  level of $175\mu m$ the bullet's signature stabilizes. For this land, matches should not be attempted at lower heights. }
\end{figure}
%An opportunity that arises from comparing multiple cross sections of a bullet's surface measurements is the following:

@@

\begin{figure}[hbtp]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.5\textwidth]{figure/smoothfac1-1}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance using the James Hamby study data. As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash' \citep{hamby:2009}.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read_x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
Ideally, results look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

\begin{figure}[hbtp]
\begin{subfigure}[t]{\textwidth}\centering
\caption{Maximal number of CMS between unknown bullet land C-3 and all of the other 118 considered (known) lands. For two lands the number of maximum CMS is high. }
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.5\textwidth]{figure/cms-1}

\end{knitrout}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Overlaid signatures of C-3 and the land with the top matching CMS.}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/top-1}

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 42 pairs of lands with at least thirteen CMS in common are matches. 
%
\begin{figure}[hbtp]
  \centering
%' \begin{subfigure}[b]{\textwidth}\centering
%' \caption{}
\begin{minipage}[t]{.47\textwidth}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/cms-bars-1}

@@

\end{knitrout}
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are 2.31 maximal CMS between known non-matches (with a standard deviation of 1.4). Known matches share on average 8.49 maximal CMS, with a standard deviation of 5.65. While the probability for a match increases with the number of maximal CMS, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch between two lands that share twelve consecutively matched striae. Visually we can easily tell that these two lands do not match well.

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between lands based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 132 lands is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the signatures is very important in the matching process  by the decision tree. Aside from  cross correlation, the total number of matches is also included in the decision rule. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2D microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitized version of the images that preserves the 3D surface structure. The findings about the discriminating power of cross-correlation are consistent with the results of the study by \citet{ma:2004}. However, in that study, the authors did not consider the number of matches and non-matches.

@@

\end{knitrout}
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is expected to be much lower for bullet-to-bullet comparisons. In the case of the Hamby data, even a single tree results in an overall error rate of zero, if we require that a match of two bullets occurs when at least two of the bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, corresponding to a rate of 0.0011. 
%\hh{XXX calling this a false negative rate is problematic. FEs will be up in arms, because inconclusive doesn't necessarily mean something was missed, it could also mean that there was something wrong with the bullets. XXX }

@@

By suggesting an automated algorithm that first removes class characteristics, such as the groove impressions, shoulders, and the curvature of the bullet to reveal the region of the land, then identifies peaks and valleys on this land, we reduce subjectivity and with it possible sources of bias. In particular, `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}. The steps outlined in this paper could also help explore other important forensic science problems. In particular, more general toolmark examination can benefit from the approach we discuss.

@@

\section*{Acknowledgment} 
Thanks to David Baldwin for pointing us to the NIST database and doing a Firearms 101 for us.
Thanks to the men and women behind the software R~\citep{R}, and the authors of the R packages  knitr~\citep{knitr} and ggplot2~\citep{ggplot2}.

>>>>

We work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}} and is implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in forensic science. We have developed an open-source package for analyzing bullet land impressions written in R \citep{R}. This package is called bulletr \citep{bulletr} and enables direct reading and manipulation of x3p files. It also implements all of the methods we propose in this paper. A different package exists for reading x3p files called x3pr \citep{x3pr}, developed by Petraco (2014), but it is not designed to carry out calculations like the ones we propose after the x3p files have been read.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[t]{\textwidth}\centering
\caption{View of the data along the circumference of the bullet (circular segment of about 30 degrees).\label{fig:sidex3p}}{%
      \includegraphics[width=.65\textwidth]{images/sidex3p.png}
      \hspace{1cm}
      \includegraphics[width=.25\textwidth]{images/side-sketch.png}
    }
\end{subfigure}    
\begin{subfigure}[t]{\textwidth}\centering
    \caption{Frontal view of a bullet land impression (lower end of the view is the bottom of the bullet). \label{fig:topx3p}}{%
    \includegraphics[width=.65\textwidth]{images/topx3p.png} \hspace{1cm}
    \includegraphics[width=.15\textwidth]{images/top-sketch.png}
    }
\end{subfigure}
\caption{Example of a groove-to-groove scan of a bullet land impression. The red-dotted rectangle on the right shows the location and orientation of the segment. }
\label{fig:allx3p}
\end{figure}

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent groove impressions on the bullet, called a ``land" or land impression. The shoulders are the raised portion of the surface between the groove impression and the land impression. In the Hamby data, typical length (shoulder-to-shoulder) of a land impression is about 1998.28 micrometers or 2 millimeters. For notational simplicity, we refer to a particular land impression of a bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these land impressions is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

\end{knitrout}
\caption{\label{fig:fixedX}Side profile of the surface measurements (in~$\mu m$) of a bullet land impression at a fixed height $x$. Note that the global features dominate any deviations, corresponding to the individual characteristics of striation marks.}
\end{figure}

@@

\subsection{Horizontal alignment}
Signatures of each of the two land impressions, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are prominent and, again, suggest a match between the land impressions. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities between signatures. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land impression signatures.\label{fig:crosscut}}{%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/crosscuts-1}

@@

\subsection{Impact of bullet height}\label{sec:heights}
The height at which signatures are extracted for a comparison between bullet land impressions matters -- signatures taken from heights that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity for quality control: we have to be aware of the height that was used in a matching. Visually, matches degrade if the signatures upon which the match is based are from heights further than 200$\mu m$ apart  (see Supplement Section~\ref{supp:ccf} for more discussion). However, we can extract signatures from multiple heights of the same bullet land impression for an initial assessment of its quality. By comparing signatures from heights that are not too far apart --  25~$\mu m$ to 50~$\mu m$ -- we get an indication whether the signatures come from a rapidly changing section of the surface, indicative of a break-off or some other damage, or from a stable section, where we have a reasonable expectation of finding matches to other signatures. In the approach here, we keep increasing the height $x$ at which the signature is taken until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet~1-1 from barrel 3, where `stability' is  defined as two aligned signatures from heights chosen 25$\mu m$ apart having a cross-correlation of at least 0.95.

@@

\end{knitrout}
\caption{\label{fig:crosscuts2}Signatures  for barrel 3, bullet~1-1 extracted from varying heights. Initially, the match between signatures taken at heights 25$\mu m$ apart is affected strongly by some break off at the bottom of the bullet. At a  level of $175\mu m$ the bullet's signature stabilizes. For this land impression, matches should not be attempted at lower heights. }
\end{figure}
%An opportunity that arises from comparing multiple cross sections of a bullet's surface measurements is the following:

@@

\begin{figure}[hbtp]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.75\textwidth]{figure/smoothfac1-1}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance using the James Hamby study data. As a first step, we automatically assess the quality of each of the land impressions by checking that we can identify a stable region. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four land impressions from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these land impressions has scratch marks across the surface, also known as `tank rash' \citep{hamby:2009}.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read_x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four land impressions that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four land impressions from further matching considerations and
 run all remaining land impressions from the unknown bullets against all remaining land impressions from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ land impressions from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ land impressions from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
Ideally, results look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land impression C-3 and all 118 land impressions from known bullets. Two land impressions show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all land impressions, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

\begin{figure}[hbtp]
\begin{subfigure}[t]{\textwidth}\centering
\caption{Maximal number of CMS between unknown bullet land impression C-3 and all of the other 118 considered (known) land impressions. For two land impressions the number of maximum CMS is high. }
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.5\textwidth]{figure/cms-1}

\end{knitrout}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Overlaid signatures of C-3 and the land impression with the top matching CMS.}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/top-1}

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 42 pairs of land impressions with at least thirteen CMS in common are matches. 
%
\begin{figure}[hbtp]
  \centering
%' \begin{subfigure}[b]{\textwidth}\centering
%' \caption{}
\begin{minipage}[t]{.47\textwidth}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/cms-bars-1}

@@

\end{knitrout}
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are 2.31 maximal CMS between known non-matches (with a standard deviation of 1.4). Known matches share on average 8.49 maximal CMS, with a standard deviation of 5.65. While the probability for a match increases with the number of maximal CMS, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch between two land impressions that share twelve consecutively matched striae. Visually we can easily tell that these two land impressions do not match well.

@@

Using recursive partitioning, we fit a decision tree~\citep{breiman:1984, rpart, rpart.plot} to predict matches between land impressions based on features derived from the image files. The resulting tree is shown in Figure~\ref{fig:tree}. A total of 132 land impressions is being matched correctly. Interestingly, the number of consecutive matching striae does not feature in this evaluation. 
Instead of CMS, cross-correlation (ccf) between the signatures is very important in the matching process  by the decision tree. Aside from  cross correlation, the total number of matches is also included in the decision rule. 
Between cross-correlation and CMS, cross-correlation has higher  predictive power. This  does not  contradict earlier findings emphasizing the value of CMS on visual assessments of bullet matches: in those papers, assessments were based on purely visual inspection of either actual bullets or 2D microscopic images of bullets.
Neither one of these methods allows for an assessment of cross-correlations. This is one of the benefits of switching to a digitized version of the images that preserves the 3D surface structure. The findings about the discriminating power of cross-correlation are consistent with the results of the study by \citet{ma:2004}. However, in that study, the authors did not consider the number of matches and non-matches.

@@

\end{knitrout}
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is expected to be much lower for bullet-to-bullet comparisons. In the case of the Hamby data, even a single tree results in an overall error rate of zero, if we require that a match of two bullets occurs when at least two of the bullet's land impressions are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, corresponding to a rate of 0.0011. 
%\hh{XXX calling this a false negative rate is problematic. FEs will be up in arms, because inconclusive doesn't necessarily mean something was missed, it could also mean that there was something wrong with the bullets. XXX }

@@

By suggesting an automated algorithm that first removes class characteristics, such as the groove impressions, shoulders, and the curvature of the bullet to reveal the region of the land impression, then identifies peaks and valleys on this land impression, we reduce subjectivity and with it possible sources of bias. In particular, `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}. The steps outlined in this paper could also help explore other important forensic science problems. In particular, more general toolmark examination can benefit from the approach we discuss.

@@

\section*{Acknowledgment} 
Thanks to David Baldwin for pointing us to the NIST database and doing a Firearms 101 for us.
Thanks to the men and women behind the software R~\citep{R}, and the authors of the R packages  knitr~\citep{knitr} and ggplot2~\citep{ggplot2}. We also wish to thank an anonymous reviewer, whose constructive and insightful comments contributed greatly to the improvement of the manuscript.
